Verifying Concurrent Search Structure Templates for C Programs using VST

### 2023-04-18 concurrency/CONCUR 2023/paper.tex
modified specs

\subsubsection{Lock-coupling Template}
\label{traverse_proof_lock}
Proceeding with the proofs, we establish the following specification for the \lstinline{traverse} function in the lock-coupling template:
%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.} 
\begin{mathpar}
	{\color{blue} {\forall \texttt{t}.\ \left\langle \nodeboxrep \texttt{(pn->n)} 
	\ast \texttt{R(pn->n)} \ | \ \treerep\ \texttt{t} \right\rangle}}\\ 
	\texttt{traverse(pn, k)} \\ 
	\color{blue} {\langle \ \texttt{res} \ \texttt{n'}. \ (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL})) \ast \nodeboxrep \texttt{(pn->n')} \ast }
	\\ 
	\color{blue} {\texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range(n')})
	\   |\ \treerep\ \texttt{t} \rangle}
\end{mathpar}

%\than{Actually, $\nodeboxrep \texttt{(pn->n)}$ and $\texttt{data\_at  n pn} \ast \texttt{my\_half g\_root} (-\inf, +\inf)$ is equivalent to $\texttt{tree\_rep}$. But now we can ignore the information of $\texttt{my\_half}$.  }

@@

As previously mentioned, the \lstinline{pn} struct includes two nodes of type \lstinline{node_t}: the $\texttt{p}$ field, signifying the parent node, and the $\texttt{n}$ field, representing the current node. We use the assertion $\ell \lockvar \texttt{R}$ to associate a resource invariant \texttt{R} with a lock location $\ell$. Thus, formally, we can express $\nodeboxrep \texttt{(pn->n)} = \texttt{pn->n->lock} \lockvar \texttt{R(pn->n)}$.

@@

\begin{align*} &\texttt{traverse\_inv(pn)} \triangleq\ \exists \texttt{n'}.\ (\texttt{k} \in \texttt{range})\land \nodeboxrep(\texttt{pn->n'})  \ast  \texttt{R(pn->n')}   
\end{align*}
The loop invariant mentioned above include specific existential variables (such as $\texttt{n'}$) that define the state of each iteration throughout the loop. The \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (see line 6 in Figure \ref{traverse_lock_b}). At this point, the specification indicates that we have $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$, enabling us to prove the post-condition. On the other hand, the function terminates if \lstinline{findNext} is unable to locate a subsequent node with the key value  $\texttt{k}$ within its set of key values (refer to line 11 in Figure \ref{traverse_lock_b}). In this case, the specification clarifies that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} \neq \texttt{NULL}$ which also enables us to prove the post-condition.

@@

\subsubsection{Give-up Template}
Moving forward with the proofs, we define the following specification for the traverse function within the give-up template:

\begin{mathpar}
	{\color{blue}{\forall \texttt{t}.\ \left\langle \infp (\texttt{pn->n}) \ | \ \treerep\ \texttt{t} \right\rangle}} \\ 
	\texttt{traverse(pn, k)} \\ 
	\color{blue}{\langle \texttt{res} \ \texttt{n'}. \ (\texttt{res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL})) \ast  \infp (\texttt{pn->n'}) \ast \texttt{R(pn->n')} \ast (\texttt{k} \in \texttt{range(n')})
	\ |\ \treerep\ \texttt{t}\rangle}
\end{mathpar}

@@

We have defined the loop invariant for the given template of the \texttt{traverse} function as follows: 
\begin{align*} &\texttt{traverse\_inv(pn)} \triangleq\ \exists \texttt{n'}.\ \infp (\texttt{pn->n}) \ast \infp (\texttt{pn->n'})   
\end{align*}

@@

\begin{figure}
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\texttt{pn->n} \ast \texttt{n->t} \ast \texttt{t} \mapsto \texttt{(k', v', l, r)} 
			\end{array}
			\right\}
		}
		\\ 
		\texttt{findNext(pn, k)} 
		\\
		{\color{blue}
			\left\{\exists \  \texttt{res n'.}
			\begin{array}{c}
				\ \texttt{pn->n} \ \ast \ \texttt{n->t} \ \ast \ \texttt{t} \mapsto \texttt{(k', v', l, r)} \  \ast \\ 
				\ \texttt{(if res then } \texttt{(l = n'} \land \texttt{k < k')} \lor \texttt{(r = n'} \land \texttt{k > k')} \texttt{ else } \texttt{(n' = n} \land \texttt{k = k'))}
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specification of \texttt{findNext} for the lock-coupling and give-up templates}
	\label{fig:findNext_lock}
\end{figure}

\begin{figure}
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\ \texttt{pn->n} \ \ast \ \texttt{n->t} 
			\end{array}
			\right\}
		}
		\\ 
		\texttt{insertOp(pn, k, v)} 
		\\
		{\color{blue}
			\left\{\exists \  \texttt{t' l r lk1 lk2}.
			\begin{array}{c}
				\ \texttt{pn->n} \ \ast \ \texttt{n->t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \ \ast \ \\ 
				\ \texttt{l} \mapsto \texttt{(NULL, lk1)}\  \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2)}
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specification of \texttt{insertOp} for the lock-coupling template}
	\label{fig:insertOp_lock}
\end{figure}

>>>>

\subsubsection{Lock-coupling Template}
\label{traverse_proof_lock}
Proceeding with the proofs, we establish the following specification for the \lstinline{traverse} function in the lock-coupling template, as illustrated in Figure \ref{fig:traverse_lock}.
%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.}

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			 \forall \  \texttt{t}. \left\langle
			\begin{array}{c}
				\nodeboxrep \texttt{(pn->n)}  \ \ast \ \texttt{R(pn->n)} \ \big| \ \treerep\ \texttt{t}
			\end{array}
			\right\rangle
		}
		\\ 
		\texttt{insertOp(pn, k, v)} 
		\\
		{\color{blue}
			\left\langle \exists \  \texttt{res} \ \texttt{n'}.
			\begin{array}{c}
				\ \texttt{(res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL))}  \ \ast \ \nodeboxrep \texttt{(pn->n')} \\  
				\ast \ \texttt{R(pn->n')} \ast \ \texttt{(k} \in \texttt{range(n'))}
			\end{array}
			\ \Bigg| \ \treerep\ \texttt{t}
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the lock-coupling template}
	\label{fig:traverse_lock}
\end{figure}
%\than{Actually, $\nodeboxrep \texttt{(pn->n)}$ and $\texttt{data\_at  n pn} \ast \texttt{my\_half g\_root} (-\inf, +\inf)$ is equivalent to $\texttt{tree\_rep}$. But now we can ignore the information of $\texttt{my\_half}$.  }

@@

As previously mentioned, the \lstinline{pn} struct includes two nodes of type \lstinline{node_t}: the $\texttt{p}$ field, signifying the parent node, and the $\texttt{n}$ field, representing the current node. We use the assertion $\ell \lockvar \texttt{R}$ to associate a resource invariant \texttt{R} with a lock location $\ell$. Thus, formally, we can express $\nodeboxrep \texttt{(pn->n)} \triangleq\ \texttt{pn->n->lock} \lockvar \texttt{R(pn->n)}$.

@@

\begin{align*} &\texttt{traverse\_inv(pn)} \triangleq\ \exists \ \texttt{n'}.\ (\texttt{k} \in \texttt{range})\ \land \ \nodeboxrep(\texttt{pn->n'})  \ \ast \  \texttt{R(pn->n')}   
\end{align*}

The loop invariant mentioned above include specific existential variables (such as $\texttt{n'}$) that define the state of each iteration throughout the loop. The \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (see line 6 in Figure \ref{traverse_lock_b}). At this point, the specification indicates that we have $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$, enabling us to prove the post-condition. On the other hand, the function terminates if \lstinline{findNext} is unable to locate a subsequent node with the key value  $\texttt{k}$ within its set of key values (refer to line 11 in Figure \ref{traverse_lock_b}). In this case, the specification clarifies that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} \neq \texttt{NULL}$ which also enables us to prove the post-condition.

@@

\subsubsection{Give-up Template}
Moving forward with the proofs, we define the following specification for the traverse function within the give-up template, as depicted in Figure \ref{fig:traverse_giveup}.

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  \texttt{t}. \left\langle
			\begin{array}{c}
				\infp (\texttt{pn->n}) \ \big| \ \treerep\ \texttt{t}
			\end{array}
			\right\rangle
		}
		\\ 
		\texttt{insertOp(pn, k, v)} 
		\\
		{\color{blue}
			\left\langle \exists \  \texttt{res} \ \texttt{n'}.
			\begin{array}{c}
				\ \texttt{(res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL))}  \ \ast \infp \texttt{(pn->n')} 
				\\ 
				\ast \ \texttt{R(pn->n')} \ \ast \ (\texttt{k} \in \texttt{range(n'))}
			\end{array}
			\ \Bigg| \ \treerep\ \texttt{t}
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the give-up template}
	\label{fig:traverse_giveup}
\end{figure}

@@

We have defined the loop invariant for the given template of the \texttt{traverse} function as follows: 
\begin{align*} &\texttt{traverse\_inv(pn)} \triangleq \ \exists \ \texttt{n'}.\ \infp (\texttt{pn->n}) \ast \infp (\texttt{pn->n'})   
\end{align*}

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\texttt{pn->n} \ast \texttt{n->t} \ast \texttt{t} \mapsto \texttt{(k', v', l, r)} 
			\end{array}
			\right\}
		}
		\\ 
		\texttt{findNext(pn, k)} 
		\\
		{\color{blue}
			\left\{\exists \  \texttt{res n'.}
			\begin{array}{c}
				\ \texttt{pn->n} \ \ast \ \texttt{n->t} \ \ast \ \texttt{t} \mapsto \texttt{(k', v', l, r)} \  \ast \\ 
				\ \texttt{(if res then } \texttt{(l = n'} \land \texttt{k < k')} \lor \texttt{(r = n'} \land \texttt{k > k')} \texttt{ else } \texttt{(n' = n} \land \texttt{k = k'))}
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specification of \texttt{findNext} for the lock-coupling and give-up templates}
	\label{fig:findNext_lock}
\end{figure}

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\ \texttt{pn->n} \ \ast \ \texttt{n->t} 
			\end{array}
			\right\}
		}
		\\ 
		\texttt{insertOp(pn, k, v)} 
		\\
		{\color{blue}
			\left\{\exists \  \texttt{t' l r lk1 lk2}.
			\begin{array}{c}
				\ \texttt{pn->n} \ \ast \ \texttt{n->t'} \ \ast \ \texttt{t'}\mapsto \texttt{(k, v, l, r)} \ \ast \ \\ 
				\ \texttt{l} \mapsto \texttt{(NULL, lk1)}\  \ast \ \texttt{r} \mapsto \texttt{(NULL, lk2)}
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specification of \texttt{insertOp} for the lock-coupling template}
	\label{fig:insertOp_lock}
\end{figure}
